Combinator Parsers: a Monad Transformer Approach

### 2013-05-08 article.tex
work on error reporting section of article

The signficance of the Maybe parameter is two-fold:  it allows computational
effects to express first success versus failure, using the Just and Nothing 
constructors, respectively, and second backtracking/alternation.  We'll
capture both of these characteristics with using a type class, Plus:
\begin{verbatim}
class Applicative f => Plus f where
  zero :: f a
  (<+>) :: f a -> f a -> f a
\end{verbatim}
This type class is identical in intent to the Alternative type class from
Control.Applicative, with zero corresponding to empty and <+> to \verb+<|>+; 
however, we aren't able to use Alternative directly because
other of the monad transformers have unsuitable instances for our purposes,
and as we'll see later, inability to swap out such instances is one of the 
limitations of the monad transformer approach.

@@

\section{Example 2: error-reporting}
While our first attempt at a parser was useful, when it encountered malformed input, it wasn't
helpful -- all it could do was throw up its hands in despair and give up.  However, we'd like
an error message that tells us what the problem was and where it was encountered whenever bad
input is encountered.  Otherwise, we'd waste time finding the problem by eye -- not a lot of 
fun.  Here are some examples of bad input that we'd like to get specific error messages for,
but currently can't:
\begin{verbatim}
unrecognized character:
application missing operator:
application missing close parentheses:
special form missing close curly:
invalid special form name:
duplicate lambda parameters:
lambda missing body forms:
lambda missing parameter list:
define missing symbol:
define missing form:
\end{verbatim}

>>>>

The significance of the Maybe parameter is two-fold:  it allows computational
effects to express first success versus failure, using the Just and Nothing 
constructors, respectively, and second backtracking/alternation.  We'll
capture both of these characteristics with using a type class, Plus:
\begin{verbatim}
class Applicative f => Plus f where
  zero :: f a
  (<+>) :: f a -> f a -> f a
\end{verbatim}
This type class is identical in intent to the Alternative type class from
Control.Applicative, with zero corresponding to empty and <+> to \verb+<|>+; 
however, we aren't able to use Alternative directly because
other of the monad transformers have unsuitable instances for our purposes,
and as we'll see later, inability to swap out such instances is one of the 
limitations of the monad transformer approach.

@@

\section{Example 2: error-reporting}
While our first attempt at a parser was useful, when it encountered malformed input, it wasn't
helpful -- all it could do was throw up its hands in despair and give up.  However, we'd like
an error message that tells us what the problem was and where it was encountered whenever bad
input is encountered.  Otherwise, we'd waste time finding the problem by eye -- not a lot of 
fun.  Here are the examples of bad input that we'd like to get specific error messages for:
\begin{itemize}
\item application: missing operator:  \begin{verbatim}()\end{verbatim}
\item application: missing close parenthesis:  \begin{verbatim}(a b\end{verbatim}
\item define: missing symbol:  \begin{verbatim}{define (a b c)}\end{verbatim}
\item define: missing form:  \begin{verbatim}{define a}\end{verbatim}
\item define: missing close curly:  \begin{verbatim}{define a b\end{verbatim}
\item lambda: missing parameter list:  \begin{verbatim}{lambda (a b c)}\end{verbatim}
\item lambda: duplicate parameter names:  \begin{verbatim}{lambda {a b a} (c d)}\end{verbatim}
\item lambda: missing parameter list close curly:  \begin{verbatim}{lambda {a b (c d)}\end{verbatim}
\item lambda: missing body form:  \begin{verbatim}{lambda {a b}}\end{verbatim}
\item lambda: missing close curly:  \begin{verbatim}{lambda {a b} (c d)\end{verbatim}
\item special form: unable to parse:  \begin{verbatim}{defin x y}\end{verbatim}
\item woof: unparsed input:  \begin{verbatim}a,b\end{verbatim}
\end{itemize}
To do this, we're going to bring in a second monad transformer type class -- MonadError -- and
another monad transformer datatype -- MaybeT.  Our parser stack and runParser function are now
going to be:
\begin{verbatim}
type Parse e t a = StateT [t] (MaybeT (Either' e)) a

runParser :: Parse e t a -> [t] -> Either' e (Maybe (a, [t]))
runParser p xs = runMaybeT (runStateT p xs)
\end{verbatim}
No, that ' isn't accidental -- we again introduce a replacement data type to avoid a type class
instance that doesn't fulfill our needs.  In this case, Either' is intended as a replacement for
Either, with the only difference being its instance of MonadError.  All other instances for 
Either' are analagous to those of Either:
\begin{verbatim}
data Either' a b
    = Left' a
    | Right' b
  deriving (Show, Eq)

instance MonadError e (Either' e) where
  throwError                =  Left'
  catchError  (Right' x) _  =  Right' x
  catchError  (Left' e)  f  =  f e
\end{verbatim}
We'll also need an instance of Switch for MaybeT:
\begin{verbatim}
instance Functor m => Switch (MaybeT m) where
  switch (MaybeT m) = MaybeT (fmap switch m)
\end{verbatim}

\subsection{Reporting position}
To be able to report the position of the error, we'll pre-process the input string, calculating
the line and column number of each character.  Then instead of our parsers operating on a list
of characters, they'll operate on a list of Tokens, where a Token is a character tupled up with
a line number and column number, which we'll express with a simple typedef and some accessor functions:
\begin{verbatim}
type Token = (Char, Int, Int)
chr  (a, _, _)  =  a
line (_, b, _)  =  b
col  (_, _, c)  =  c
\end{verbatim}
The function responsible for calculating the position will simply transform a list of characters into
a list of tokens:
\begin{verbatim}
countLineCol :: [Char] -> [Token]
countLineCol = reverse . snd . foldl f ((1, 1), [])
  where
    f ((line, col), ts) '\n' = ((line + 1, 1), ('\n', line, col):ts)
    f ((line, col), ts)  c   = ((line, col + 1), (c, line, col):ts)
\end{verbatim}

\subsection{Token parsers}
Unfortunately, by adding the position to the input, we've created a new problem for ourselves: our 
original token parsers assumed they would be operating on character lists, so we'll have to rewrite
them to get rid of that assumption.  The cause of the problem is the \verb+literal+ combinator, which
we can easily replace with this one that assumes token lists:
\begin{verbatim}
character :: (MonadState [Token] m, Plus m) => Char -> m Token
character c = satisfy ((==) c . chr)
\end{verbatim}
And now our token parsers are:
\begin{verbatim}
whitespace = many1 $ satisfy (flip elem " \n\t\r\f" . chr)
comment = pure (:) <*> character ';' <*> many0 (not1 $ character '\n')

munch p = many0 (whitespace <+> comment) *> p

ocurly = munch $ character '{'
ccurly = munch $ character '}'
oparen = munch $ character '('
cparen = munch $ character ')'
symbol = munch $ many1 char
  where char = fmap chr $ satisfy (flip elem (['a' .. 'z'] ++ ['A' .. 'Z']) . chr)
\end{verbatim}
The only differences are that literal was replaced everywhere by character, and a couple uses
of the chr accessor.  The munch combinator didn't have to change at all (although its inferred
type did in fact change).

\subsection{Error messages}
Now we get to the meat of this section -- modifying the parsers to report accurate and useful errors.
Let's start out with a list, in code, of all the errors that we're going to deal with:
\begin{verbatim}
eaOp   =  "application: missing operator"
eaCls  =  "application: missing close parenthesis"
edSym  =  "define: missing symbol"
edForm =  "define: missing form"
edCls  =  "define: missing close curly"
elPL   =  "lambda: missing parameter list"
elPrms =  "lambda: duplicate parameter names"
elPCls =  "lambda: missing parameter list close curly"
elBody =  "lambda: missing body form"
elCls  =  "lambda: missing close curly"
esName =  "special form: unable to parse"
ewUnp  =  "woof: unparsed input"
\end{verbatim}
The MonadError type class provides two useful combinators, throwError for generating errors, and 
catchError for dealing with them.  We can ignore catchError because we don't want to try to recover
from errors, we just want them reported.  Now the question is, when should an error be created
and what data should it include?  \verb+throwError+ is too basic for our needs, so we'll introduce
the \verb+commit+ combinator:
\begin{verbatim}
commit :: (MonadError e m, Plus m) => e -> m a -> m a
commit err p = p <+> throwError err
\end{verbatim}
This combinator takes two arguments: 1. an error value, and 2. a monadic computation.  It first tries
to run the computation, and if the computation fails, generates an error whose contents are the first
argument.  The definition of 'failure' is provided by the \verb+Plus+ instance of \verb+m+.
So what data do we want to include in our errors?  A String describing the error and a token reporting
the position should describe the error decently:
\begin{verbatim}
type Error = (String, Token)
\end{verbatim}
Now lets start with \verb+Application+ as our first modified, error-reporting parser:
\begin{verbatim}
application =
    oparen                       >>= \open ->
    commit (eaOp, open) form     >>= \op ->
    many0 form                   >>= \args ->
    commit (eaCls, open) cparen  >>
    return (AApp op args)
\end{verbatim}
This parser works by running the open-parenthesis parser, capturing its value in \verb+open+, and continuing
with the parsing; if either \verb+form+ or close-parenthesis fails, an appropriate error can be generated
which includes a string and the location of the open parenthesis.  However, if the open parenthesis is not
successfully parsed, it's a failure but not an error.  Here are some examples to demonstrate this in action:
\begin{verbatim}
ghci$ runParser application (countLineCol "oops")
Right' Nothing

ghci$ runParser application (countLineCol "(a b c)")
Right' (Just (AApp (ASymbol "a") [ASymbol "b",ASymbol "c"],[]))

ghci$ runParser application (countLineCol "\n()")
Left' ("application: missing operator",('(',2,1))

ghci$ runParser application (countLineCol "(a b")
Left' ("application: missing close parenthesis",('(',1,1))
\end{verbatim}
Although the previous version of the \verb+application+ parser could be implemented with just applicative combinators, 
this version certainly can't:  \verb+open+ is used to build the following parsers, making this decidedly monadic.

The \verb+define+ and \verb+lambda+ parsers are upgraded similarly:
\begin{verbatim}
define =
    ocurly                       >>= \open ->
    check (== "define") symbol    *>
    pure ADefine                 <*>
    commit (edSym, open) symbol  <*>
    commit (edForm, open) form   <*
    commit (edCls, open) ccurly

lambda =
    ocurly                               >>= \open ->
    check (== "lambda") symbol           >>
    commit (elPL, open) ocurly           >>= \p_open ->
    many0 symbol                         >>= \params ->
    (if distinct params 
        then return ()
        else throwError (elPrms, open))  >>
    commit (elPCls, open) ccurly         >>
    commit (elBody, open) (many1 form)   >>= \bodies ->
    commit (elCls, open) ccurly          >>
    return (ALambda params bodies)
  where
    distinct names = length names == length (nub names)
\end{verbatim}
Note that \verb+commit+ isn't used until we can be sure that we're in the right rule:  if, in the \verb+define+
parser, we committed to parsing the "define" keyword, we wouldn't be able to backtrack and try the \verb+lambda+
parser if that failed.  Thus, if treating errors as unrecoverable, it's important not to place a \verb+commit+
where backtracking might be necessary to parse a valid alternative.  However, this isn't the case with how
we've used it, since once we see an open curly brace and a define, we're sure that we're in the \verb+define+
rule, and allowing backtracking would only destroy our ability to produce clean and accurate error reports.

\subsection{Error messages:  finishing touches}
To complete the error-reporting parser, we need to change a couple more details.  First, we want an error 
reported when an open curly brace is found, but no special form can be reported.  To accomplish this, we
can extend the \verb+special+ parser with a third alternative, which results in an error if it can parse
an open curly brace, but in a backtracking failure if not:
\begin{verbatim}
special = define <+> lambda <+> (ocurly >>= \o -> throwError (esName, o))
\end{verbatim}
The \verb+form+ parser doesn't have to be changed:
\begin{verbatim}
form = fmap ASymbol symbol <+> application <+> special
\end{verbatim}
When we can no longer parse any forms, instead of failing, we'd like to report an error if there's any input
left, and otherwise succeed.  To implement this parser, we'll first munch any leading whitespace or comments, 
as usual, and then we'll check the input:  if it's empty, great; if not, report an error at the location of the
first remaining token, along with the appropriate error message that we defined earlier:
\begin{verbatim}
endCheck = 
    many0 (whitespace <+> comment) *>
    get >>= \xs -> case xs of
                        (t:_) -> throwError (ewUnp, t)
                        []    -> pure ()
\end{verbatim}
Now our \verb+woof+ parser uses this \verb+endCheck+ parser after it has consumed all the forms that it can:
\begin{verbatim}
woof = many0 form <* endCheck
\end{verbatim}

\subsection{Error messages:  examples}
While correct input will be parsed in the same way by our new parsers, incorrect input will now (hopefully) be
accurately reported as such.  Successful results will be slightly different, however -- there is now an extra
constructor wrapper around the value:
\begin{verbatim}
runParser woof (countLineCol "{define fgh {lambda {f x y} {define a (b c)} (f y a (a x))}}")
Right' (Just ([ADefine "fgh" 
                       (ALambda ["f","x","y"] 
                                [ADefine "a" (AApp (ASymbol "b") [ASymbol "c"]),
                                 AApp (ASymbol "f") [ASymbol "y",
                                                     ASymbol "a",
                                                     AApp (ASymbol "a") [ASymbol "x"]]])],
              []))
\end{verbatim}
Here are examples showing errors detected by the parser:
\begin{verbatim}
ghci$ map (runParser woof . countLineCol) ["()", "(a b", "{define (a b c)}",  
                                     "{define a}", "{define a b", 
                                     "{lambda (a b c)}", "{lambda {a b a} (c d)}", 
                                     "{lambda {a b (c d)}", "{lambda {a b}}", 
                                     "{lambda {a b} (c d)", "{defin x y}", "a,b"]
Left' ("application: missing operator",('(',1,1))
Left' ("application: missing close parenthesis",('(',1,1))
Left' ("define: missing symbol",('{',1,1))
Left' ("define: missing form",('{',1,1))
Left' ("define: missing close curly",('{',1,1))
Left' ("lambda: missing parameter list",('{',1,1))
Left' ("lambda: duplicate parameter names",('{',1,1))
Left' ("lambda: missing parameter list close curly",('{',1,1))
Left' ("lambda: missing body form",('{',1,1))
Left' ("lambda: missing close curly",('{',1,1))
Left' ("special form: unable to parse",('{',1,1))
Left' ("woof: unparsed input",(',',1,2))
\end{verbatim}
